Label object representation and manipulation with ITK

### 2007-09-24 article/Article.tex
minor corrections
darcs-hash:20070924101624-2fc9d-c9b5fca7687c82ce82841bee9f642f07a7fc9b80.gz

The classes {\em itk::ShapeLabelObject} and {\em itk::StatisticsLabelObject} have been
created to reduce the number of filters made to manipulate the attributes, and to make
the computation of all the set of attributes much efficient. In the early stage of
development, all the attributes were managed as in {\em AttributeLabelObject}, and
a set of 8 classes made to manipulate a single attribute were provided, leading to a huge
number of classes.

@@

\begin{itemize}
  \item {\em Size} is the size of the object in number of pixels. Its type is
{\em unsigned long}.
  \item {\em PhysicalSize} is the size of the object in physical unit. It is equal
to the {\em Size} multiplicated by the physical pixel size. Its type is {\em double}.
  \item {\em Centroid} is the position of the center of the shape in physical coordinates. It is not
constrained to be in the object, and thus can be outside if the object is not convex.
Its type is {\em Point< double, ImageDimension >}. 
  \item {\em Region} is the bounding box of the object given in the pixel coordinates.
The physical coordinate can easily be computed from it. Its type is
{\em ImageRegion< ImageDimension >}.
  \item {\em RegionElongation} is the ratio of the longest physical size of the region
on one dimension and its smallest physical size. This descriptor is not robust, and in
particular is sensitive to rotation. Its type is {\em double}.
  \item {\em SizeRegionRatio} is the ratio of the size of the object region (the
bounding box) and the real size of the object. Its type is {\em double}.
  \item {\em SizeOnBorder} is the number of pixels in the objects which are on the border
of the image. A pixel on several borders (a pixel in a corner) is counted only one time, so the
size on border can't be greater than the size of the object.
This attribute is particulary useful to remove the objects which are touching
too much the border. Its type is {\em unsigned long}.
  \item {\em PhysicalSizeOnBorder} is the physical size of the objects which are on the border
of the image. In 2D, it is a distance, in 3D, a surface, etc. Contrary to the {\em PhysicalSize}
attribute which is directly linked to the {\em Size}, this attribute is not directly
linked to the {\em SizeOnBorder} attribute. This attribute is particulary useful to remove
the objects which are touching too much the border. Its type is {\em double}.
  \item {\em FeretDiameter} is the diameter in physical units of the sphere which include
all the object. The feret diameter is not computed by default, because of its high computation.
Its type is {\em double}.
  \item {\em BinaryPrincipalMoments} contains the principal moments. Its type is
{\em itk::Vector< double, ImageDimension >}.
  \item {\em BinaryPrincipalAxes} contains the principal axes of the object. Its type is
{\em itk::Matrix< double, ImageDimension, ImageDimension >}.
  \item {\em BinaryElongation} is the elongation of the shape, computed as the ratio
of the largest principal moment by the smallest principal moment. Its value is greater
or equal to $1$. Its type si {\em double}.
%   \item {\em Perimeter}
\end{itemize}

@@

\begin{itemize}
  \item {\em Minimum} is the minimum value in the feature image for the object. Its type is
the feature image pixel type.
  \item {\em MinimumIndex} is the index position in the image where the first minimum was
found. Its type is{\em Index< ImageDimension >}.
  \item {\em Maximum} is the maximum value in the feature image for the object. Its type is
the feature image pixel type.
  \item {\em MaximumIndex} is the index position in the image where the first maximum was
found. Its type is{\em Index< ImageDimension >}.
  \item {\em Mean} is the mean of the pixel values in the object. Its type is {\em double}.
  \item {\em Sum} is the sum of all the pixel values in the objects. Its type is {\em double}.
  \item {\em Sigma} is the standard deviation of the pixels values in the objects. Its type
is {\em double}.
  \item {\em Variance} is the variance of the pixels values in the objects. Its type is
{\em double}.
  \item {\em Median} is the median of the pixels values in the obejct. Its type is
{\em double}
  \item {\em CenterOfGravity} is the center of gravity of the object. It type is
{\em Point< double >}.
  \item {\em Kurtosis} is the kurtosis of the pixel values in the objects. Its type is
{\em double}.
  \item {\em Skewness} is the skewness of the pixel values in the objects. Its type is
{\em double}.
  \item {\em PrincipalMoments} contains the principal moments. Its type is
{\em itk::Vector< double, ImageDimension >}.
  \item {\em PrincipalAxes} contains the principal axes of the object. Its type is
{\em itk::Matrix< double, ImageDimension, ImageDimension >}.
  \item {\em Elongation} is the elongation of the shape, computed as the ratio
of the largest principal moment by the smallest principal moment. Its value is greater
or equal to $1$· Its type si {\em double}.
\end{itemize}

@@

The {\em AttributeLabelObject} let the user specify the type of the attribute
he wants to use, and thus is the good choice to implement a new attribute.

@@

\end{verbatim} \normalsize
then we can read the attribute values we're interested in. {\em BinaryImageToLabelMapFilter}
produces consecutives labels, so a simple {for} loop will do the job.
\small \begin{verbatim}
  LabelCollectionType::Pointer collection = shape->GetOutput();
  for( int label=1; label<collection->GetNumberOfLabelObjects(); label++ )
    {
    LabelObjectType::Pointer labelObject = collection->GetLabelObject( label );
    std::cout << label << "\t" << labelObject->GetPhysicalSize() << "\t" << labelObject->GetCentroid() << std::endl;
    }

@@

Lets begin with the usual {\em imports}.
\small \begin{verbatim}
import itk, sys
itk.auto_progress()
\end{verbatim} \normalsize
Then declare the type we will use, as is {\em c++}.
\small \begin{verbatim}
Dimension = 2
PixelType = itk.UC
ImageType = itk.Image[ PixelType, Dimension ]

@@

LabelObjectType = itk.StatisticsLabelObject[itk.UL, Dimension]
LabelMapType = itk.LabelMap[LabelObjectType]
\end{verbatim} \normalsize
read the image of the nucleus
\small \begin{verbatim}
nuclei = itk.ImageFileReader[ImageType].New(FileName="images/noyaux.png")
\end{verbatim} \normalsize
perform a simple binarization. Note that the Otsu filter does not use the same convention as usual: the white part is outside.
\small \begin{verbatim}
otsu = itk.OtsuThresholdImageFilter[ImageType, ImageType].New(nuclei,
    OutsideValue=255, InsideValue=0)
\end{verbatim} \normalsize
The nuclei are not separated. We split them with a watershed.
\small \begin{verbatim}
maurer = itk.SignedMaurerDistanceMapImageFilter[ImageType, DistanceImageType].New(otsu)
watershed = itk.MorphologicalWatershedImageFilter[DistanceImageType, ImageType].New(maurer,
    Level=60, MarkWatershedLine=False)
mask = itk.MaskImageFilter[ImageType, ImageType, ImageType].New(watershed, otsu)
\end{verbatim} \normalsize
And now switch to the label collection representation
\small \begin{verbatim}
label = itk.LabelImageToLabelMapFilter[ImageType, LabelMapType]
    .New(mask)
\end{verbatim} \normalsize
compute the attribute values
\small \begin{verbatim}
stats = itk.StatisticsLabelMapFilter[LabelMapType, ImageType]
    .New(label, nuclei)
\end{verbatim} \normalsize
drop the objects too small to be a nucleus, and the ones on the border
\small \begin{verbatim}
size = itk.ShapeOpeningLabelMapFilter[LabelMapType].New(stats,
    Attribute='Size', Lambda=100)
border = itk.ShapeOpeningLabelMapFilter[LabelMapType].New(size,
    Attribute='SizeOnBorder', Lambda=10, ReverseOrdering=True)
\end{verbatim} \normalsize
Reoder the labels. The objects with the highest mean are the first ones.
\small \begin{verbatim}
relabel = itk.StatisticsRelabelLabelMapFilter[LabelMapType]
    .New(border, Attribute='Mean')
\end{verbatim} \normalsize
for visual validation:
\small \begin{verbatim}
labelNuclei = itk.LabelMapToLabelImageFilter[LabelMapType,
    ImageType].New(relabel)
overlay = itk.LabelOverlayImageFilter[ImageType, ImageType, RGBImageType].New(nuclei,
    labelNuclei, UseBackground=True)
itk.write(overlay, "nuclei-overlay.png")
\end{verbatim} \normalsize

@@

Now, the spots:
\small \begin{verbatim}
spots = itk.ImageFileReader[ImageType].New(FileName="images/spots.png")
\end{verbatim} \normalsize
Mask the spot image to keep only the nucleus zone. The rest of the image is cropped, excepted a border of 2 pixels
\small \begin{verbatim}
maskSpots = itk.LabelMapMaskImageFilter[LabelMapType, ImageType].New(relabel, spots, Label=1,
    Crop=True, CropBorder=2)
\end{verbatim} \normalsize
A simple thresholding:
\small \begin{verbatim}
th = itk.BinaryThresholdImageFilter[ImageType, ImageType].New(maskSpots, LowerThreshold=110)
\end{verbatim} \normalsize
Now swith to the label collection representation, and compute the attribute values. This time, the input image is not a label image, but a binary one.
\small \begin{verbatim}
slabel = itk.BinaryImageToLabelMapFilter[ImageType, LabelMapType]
    .New(th)
sstats = itk.StatisticsLabelMapFilter[LabelMapType, ImageType]
    .New(slabel, nuclei)
\end{verbatim} \normalsize
we know there are for spots in the nubleus, so keep the 4 biggest spots. The other attribute are also usable - we may have chosen to keep the 4 brightest spots for example.
\small \begin{verbatim}
skeep = itk.ShapeKeepNObjectsLabelMapFilter[LabelMapType].New(
    sstats, Attribute='Size', NumberOfObjects=4)
\end{verbatim} \normalsize
Reoder the labels. The bigger objects first.
\small \begin{verbatim}
srelabel = itk.StatisticsRelabelLabelMapFilter[LabelMapType].New(
    skeep, Attribute='Size')
\end{verbatim} \normalsize
Finally, display the values we are interested in:
\begin{itemize}
  \item the nucleus number,
  \item the spot position,
  \item the mean value in the nucleus in the spot zone.
\end{itemize}

@@

The {\em BinaryImageToLabelMapFilter} class is a slight modification of the
Richard Beare's {\em ConnectedComponentImageFilter}, and thus, has not been threaded.
It should however be possible to increase its performance that way.

The classical thread architecture is used when the input image is an {\em Image}: the image
is splitted in several regions (one per thread), and each thread work on its own region.

Because the {\em LabelMap} image is not an array of pixels, it can't be splitted
that way. Instead, several threads are created, and try to take an object in the collection.
If they get one, they process that object individually, and try to get another one when the
object is processed. If no object can be get, the thread ends. A {\em FastMutexLock} is used
to ensure that only one thread take an object at a time.

For the developer, the usage of the threading support is made very simple, by subclassing
{\em LabelMapFilter}, or {\em InPlaceLabelMapFilter}, and implementing
the method {\em virtual void ThreadedGenerateData( LabelObjectType * labelObject )} in the
new class. This method only has to process the labelObject passed in parameter. All the
threading code and mutex lock management is already implemented. The mutex lock remain
accessible if the subclass need to use it, as the {\em m\_LabelObjectContainerLock} ivar.

@@

All the filters which are taking a {\em LabelMap} as input, and are producing a {\em LabelMap} as output, are implemented as a subclass of {\em InPlaceLabelMapFilter} and
thus are running in place by default.

@@

It may be useful to implement the most commonly used opening, keep N objects and relabel transforms in a more efficient way, by using an {\em AttributeLabelObject} instead of a {\em ShapeLabelObject} or a {\em StatisticsLabelObject}.

The {\em BinaryImageToLabelMapFilter} class should be threaded to get the best of that filter on multiprocessors systems.

@@

Finally, all the binary and label filters should be implemented as a subclass of {\em InPlaceImageFilter}.

@@

\section{Acknowledgments}
I thank Richard Beare for his suggestion to use the run length encoding to represent the binary objects, and Julien Jomier for his help for the choice to {\em not} use the {\em itk::SpatialObject} class as base class of the {\em itk::LabelObejct} class.

>>>>

The classes {\em itk::ShapeLabelObject} and {\em itk::StatisticsLabelObject} have been
created to reduce the number of filters made to manipulate the attributes, and to make
the computation of all the set of attributes much efficient. In the early stage of
development, all the attributes were managed as in {\em itk::AttributeLabelObject}, and
a set of 8 classes made to manipulate a single attribute were provided, leading to a huge
number of classes.

@@

\begin{itemize}
  \item {\em Size} is the size of the object in number of pixels. Its type is
{\em unsigned long}.
  \item {\em PhysicalSize} is the size of the object in physical unit. It is equal
to the {\em Size} multiplicated by the physical pixel size. Its type is {\em double}.
  \item {\em Centroid} is the position of the center of the shape in physical coordinates. It is not
constrained to be in the object, and thus can be outside if the object is not convex.
Its type is {\em itk::Point< double, ImageDimension >}. 
  \item {\em Region} is the bounding box of the object given in the pixel coordinates.
The physical coordinate can easily be computed from it. Its type is
{\em itk::ImageRegion< ImageDimension >}.
  \item {\em RegionElongation} is the ratio of the longest physical size of the region
on one dimension and its smallest physical size. This descriptor is not robust, and in
particular is sensitive to rotation. Its type is {\em double}.
  \item {\em SizeRegionRatio} is the ratio of the size of the object region (the
bounding box) and the real size of the object. Its type is {\em double}.
  \item {\em SizeOnBorder} is the number of pixels in the objects which are on the border
of the image. A pixel on several borders (a pixel in a corner) is counted only one time, so the
size on border can't be greater than the size of the object.
This attribute is particulary useful to remove the objects which are touching
too much the border. Its type is {\em unsigned long}.
  \item {\em PhysicalSizeOnBorder} is the physical size of the objects which are on the border
of the image. In 2D, it is a distance, in 3D, a surface, etc. Contrary to the {\em PhysicalSize}
attribute which is directly linked to the {\em Size}, this attribute is not directly
linked to the {\em SizeOnBorder} attribute. This attribute is particulary useful to remove
the objects which are touching too much the border. Its type is {\em double}.
  \item {\em FeretDiameter} is the diameter in physical units of the sphere which include
all the object. The feret diameter is not computed by default, because of its high computation.
Its type is {\em double}.
  \item {\em BinaryPrincipalMoments} contains the principal moments. Its type is
{\em itk::Vector< double, ImageDimension >}.
  \item {\em BinaryPrincipalAxes} contains the principal axes of the object. Its type is
{\em itk::Matrix< double, ImageDimension, ImageDimension >}.
  \item {\em BinaryElongation} is the elongation of the shape, computed as the ratio
of the largest principal moment by the smallest principal moment. Its value is greater
or equal to $1$. Its type si {\em double}.
%   \item {\em Perimeter}
\end{itemize}

@@

\begin{itemize}
  \item {\em Minimum} is the minimum value in the feature image for the object. Its type is
the feature image pixel type.
  \item {\em MinimumIndex} is the index position in the image where the first minimum was
found. Its type is{\em itk::Index< ImageDimension >}.
  \item {\em Maximum} is the maximum value in the feature image for the object. Its type is
the feature image pixel type.
  \item {\em MaximumIndex} is the index position in the image where the first maximum was
found. Its type is{\em itk::Index< ImageDimension >}.
  \item {\em Mean} is the mean of the pixel values in the object. Its type is {\em double}.
  \item {\em Sum} is the sum of all the pixel values in the objects. Its type is {\em double}.
  \item {\em Sigma} is the standard deviation of the pixels values in the objects. Its type
is {\em double}.
  \item {\em Variance} is the variance of the pixels values in the objects. Its type is
{\em double}.
  \item {\em Median} is the median of the pixels values in the obejct. Its type is
{\em double}
  \item {\em CenterOfGravity} is the center of gravity of the object. It type is
{\em itk::Point< double >}.
  \item {\em Kurtosis} is the kurtosis of the pixel values in the objects. Its type is
{\em double}.
  \item {\em Skewness} is the skewness of the pixel values in the objects. Its type is
{\em double}.
  \item {\em PrincipalMoments} contains the principal moments. Its type is
{\em itk::Vector< double, ImageDimension >}.
  \item {\em PrincipalAxes} contains the principal axes of the object. Its type is
{\em itk::Matrix< double, ImageDimension, ImageDimension >}.
  \item {\em Elongation} is the elongation of the shape, computed as the ratio
of the largest principal moment by the smallest principal moment. Its value is greater
or equal to $1$· Its type si {\em double}.
\end{itemize}

@@

The {\em itk::AttributeLabelObject} let the user specify the type of the attribute
he wants to use, and thus is the good choice to implement a new attribute.

@@

\end{verbatim} \normalsize
then we can read the attribute values we're interested in. {\em itk::BinaryImageToLabelMapFilter}
produces consecutives labels, so a simple {for} loop will do the job.
\small \begin{verbatim}
  LabelCollectionType::Pointer collection = shape->GetOutput();
  for( int label=1; label<collection->GetNumberOfLabelObjects(); label++ )
    {
    LabelObjectType::Pointer labelObject = collection->GetLabelObject( label );
    std::cout << label << "\t" << labelObject->GetPhysicalSize() << "\t" << labelObject->GetCentroid() << std::endl;
    }

@@

Lets begin with the usual {\em import}s.
\small \begin{verbatim}
import itk, sys
itk.auto_progress()
\end{verbatim} \normalsize
Then declare the type we will use, as is {\em c++}.
\small \begin{verbatim}
Dimension = 2
PixelType = itk.UC
ImageType = itk.Image[ PixelType, Dimension ]

@@

LabelObjectType = itk.StatisticsLabelObject[itk.UL, Dimension]
LabelMapType = itk.LabelMap[LabelObjectType]
\end{verbatim} \normalsize
read the image of the nucleus
\small \begin{verbatim}
nuclei = itk.ImageFileReader[ImageType].New(FileName="images/noyaux.png")
\end{verbatim} \normalsize
perform a simple binarization. Note that the Otsu filter does not use the same convention as usual: the white part is outside.
\small \begin{verbatim}
otsu = itk.OtsuThresholdImageFilter[ImageType, ImageType].New(nuclei, OutsideValue=255,
    InsideValue=0)
\end{verbatim} \normalsize
The nuclei are not separated. We split them with a watershed.
\small \begin{verbatim}
maurer = itk.SignedMaurerDistanceMapImageFilter[ImageType, DistanceImageType].New(otsu)
watershed = itk.MorphologicalWatershedImageFilter[DistanceImageType, ImageType].New(maurer,
    Level=60, MarkWatershedLine=False)
mask = itk.MaskImageFilter[ImageType, ImageType, ImageType].New(watershed, otsu)
\end{verbatim} \normalsize
And now switch to the label collection representation
\small \begin{verbatim}
label = itk.LabelImageToLabelMapFilter[ImageType, LabelMapType].New(mask)
\end{verbatim} \normalsize
compute the attribute values
\small \begin{verbatim}
stats = itk.StatisticsLabelMapFilter[LabelMapType, ImageType].New(label, nuclei)
\end{verbatim} \normalsize
drop the objects too small to be a nucleus, and the ones on the border
\small \begin{verbatim}
size = itk.ShapeOpeningLabelMapFilter[LabelMapType].New(stats,Attribute='Size',
    Lambda=100)
border = itk.ShapeOpeningLabelMapFilter[LabelMapType].New(size, Attribute='SizeOnBorder',
    Lambda=10, ReverseOrdering=True)
\end{verbatim} \normalsize
Reoder the labels. The objects with the highest mean are the first ones.
\small \begin{verbatim}
relabel = itk.StatisticsRelabelLabelMapFilter[LabelMapType].New(border, Attribute='Mean')
\end{verbatim} \normalsize
for visual validation:
\small \begin{verbatim}
labelNuclei = itk.LabelMapToLabelImageFilter[LabelMapType, ImageType].New(relabel)
overlay = itk.LabelOverlayImageFilter[ImageType, ImageType, RGBImageType].New(nuclei,
    labelNuclei, UseBackground=True)
itk.write(overlay, "nuclei-overlay.png")
\end{verbatim} \normalsize

@@

Now, the spots:
\small \begin{verbatim}
spots = itk.ImageFileReader[ImageType].New(FileName="images/spots.png")
\end{verbatim} \normalsize
Mask the spot image to keep only the nucleus zone. The rest of the image is cropped, excepted a border of 2 pixels
\small \begin{verbatim}
maskSpots = itk.LabelMapMaskImageFilter[LabelMapType, ImageType].New(relabel, spots, Label=1,
    Crop=True, CropBorder=2)
\end{verbatim} \normalsize
A simple thresholding:
\small \begin{verbatim}
th = itk.BinaryThresholdImageFilter[ImageType, ImageType].New(maskSpots, LowerThreshold=110)
\end{verbatim} \normalsize
Now swith to the label collection representation, and compute the attribute values. This time, the input image is not a label image, but a binary one.
\small \begin{verbatim}
slabel = itk.BinaryImageToLabelMapFilter[ImageType, LabelMapType].New(th)
sstats = itk.StatisticsLabelMapFilter[LabelMapType, ImageType].New(slabel, nuclei)
\end{verbatim} \normalsize
we know there are for spots in the nubleus, so keep the 4 biggest spots. The other attribute are also usable - we may have chosen to keep the 4 brightest spots for example.
\small \begin{verbatim}
skeep = itk.ShapeKeepNObjectsLabelMapFilter[LabelMapType].New(sstats, Attribute='Size',
    NumberOfObjects=4)
\end{verbatim} \normalsize
Reoder the labels. The bigger objects first.
\small \begin{verbatim}
srelabel = itk.StatisticsRelabelLabelMapFilter[LabelMapType].New(skeep, Attribute='Size')
\end{verbatim} \normalsize
Finally, display the values we are interested in:
\begin{itemize}
  \item the nucleus number,
  \item the spot position,
  \item the mean value in the nucleus in the spot zone.
\end{itemize}

@@

The {\em itk::BinaryImageToLabelMapFilter} class is a slight modification of the
Richard Beare's {\em itk::ConnectedComponentImageFilter}, and thus, has not been threaded.
It should however be possible to increase its performance that way.

The classical thread architecture is used when the input image is an {\em itk::Image}: the image
is splitted in several regions (one per thread), and each thread work on its own region.

Because the {\em itk::LabelMap} image is not an array of pixels, it can't be splitted
that way. Instead, several threads are created, and try to take an object in the collection.
If they get one, they process that object individually, and try to get another one when the
object is processed. If no object can be get, the thread ends. A {\em itk::FastMutexLock} is used
to ensure that only one thread take an object at a time.

For the developer, the usage of the threading support is made very simple, by subclassing
{\em itk::LabelMapFilter}, or {\em itk::InPlaceLabelMapFilter}, and implementing
the method {\em virtual void ThreadedGenerateData( LabelObjectType * labelObject )} in the
new class. This method only has to process the labelObject passed in parameter. All the
threading code and mutex lock management is already implemented. The mutex lock remain
accessible if the subclass need to use it, as the {\em m\_LabelObjectContainerLock} ivar.

@@

All the filters which are taking a {\em itk::LabelMap} as input, and are producing a {\em itk::LabelMap} as output, are implemented as a subclass of {\em InPlaceLabelMapFilter} and
thus are running in place by default.

@@

It may be useful to implement the most commonly used opening, keep N objects and relabel transforms in a more efficient way, by using an {\em itk::AttributeLabelObject} instead of a {\em itk::ShapeLabelObject} or a {\em itk::StatisticsLabelObject}.

The {\em itk::BinaryImageToLabelMapFilter} class should be threaded to get the best of that filter on multiprocessors systems.

@@

Finally, all the binary and label filters should be implemented as a subclass of {\em itk::InPlaceImageFilter}.

@@

\section{Acknowledgments}
I thank Richard Beare for his suggestion to use the run length encoding to represent the binary objects, and Julien Jomier for his help for the choice to {\em not} use the {\em itk::SpatialObject} class as base class of the {\em itk::LabelObject} class.
